Compositional Certified Resource Bounds

### 2014-07-03 paper/main.tex
updates to intro

\category{D.2.4}{Software Engineering}
{Software/Program Verification}
\category{F.3.1}{Logics and Meanings of Programs}
{Specifying and Verifying and Reasoning about Programs}

\terms Verification, Reliability

\keywords Formal Verification, Compiler Construction, Program Logics,
Stack-Space Bounds, Quantitative Verification

@@

  abstract interpretation based invariant

@@

  analysis works precisely and efficiently for common loop and

@@

Our starting point is a recent work~\cite{veristack14} in which we
have formally verified bounds on the stack usage of C programs.  A key
part of this stack-bound verification is based on a quantitative Hoare
logic for deriving abstract stack bounds for CompCert Clight programs
that depend on the depth of (recursive) function calls.  % For
% non-recursive programs it is possible to compute derivations for such
% stack bounds in the quantitative Hoare logic automatically.  
By compiling these bounds together with the C program, we have derived
verified stack bounds for x86 assembly code.

Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  The
quantitative logic is implemented in Coq and proved sound with respect
to a small-step operational cost semantics for Clight that counts a
constant resource consumption for each atomic execution step.  The
resource consumption at each step is parametric in a user-definable cost
metric and can be negative to express the release of resources (as
needed, e.g., to reason about memory usage).  Furthermore, a specific
\emph{tick} function can be inserted in the source code to further
customize the cost model.  In the implementation of the logic, we use
a shallow embedding in Coq which makes the logic very flexible.  There
is practically no limitation on the format of the derived resource
bounds since they can be any function that is definable in Coq.  The
derived bounds can also be parametric in a set of cost metrics or
specifically apply to one fixed cost metric.

The precision and expressivity of the quantitative Hoare logic provide
an ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the derived
\emph{quantitative Hoare triples} can be used to specify the resource
consumption of library functions or to describe resource contracts
between different functions.  However, reasoning about quantitative
properties can often be more tedious than reasoning about intentional
properties.  Consequently, automation is inevitable to derive resource
bounds for large code bases like CertiKOS.  Such an automation was
easily achievable in the case of constant stack bounds for a code base
that does not contain recursive functions~\cite{veristack14}.  In
general however, resource bounds have to be parametric in the
arguments of a function and depend on the number of loop iterations
and (recursive) function calls that are performed by the function.

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been our main inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  However, it is a well-known problem that
type-based amortized resource analysis works well for functional
programs with pattern matching but not for C-like programs with
control flow that depends integer arithmetic.  For example, it is a
long-time open problem how to use type-based amortized analysis to
derive bounds for functions whose resource consumption depends on
(possibly negative) integers and non-sequential control flow as
introduced by \code{break} and \code{return} statements.
Unfortunately, these are the functions with non-constant resource
consumption that are most common in CertiKOS and other system
software.

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive functions as well as
\code{break} and \code{return} statements.  It targets programs that
are common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%
The use of interval sizes in potential functions (an therefore in
resource bounds) is the main innovation that makes the analysis
practical for C.

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract-interpretation based methods.  In contrast with abstract-interpretation
based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like
memory that may become available during the execution.

>>>>

% \category{D.2.4}{Software Engineering}
% {Software/Program Verification}
% \category{F.3.1}{Logics and Meanings of Programs}
% {Specifying and Verifying and Reasoning about Programs}

% \terms Verification, Reliability

% \keywords Formal Verification, Compiler Construction, Program Logics,
% Stack-Space Bounds, Quantitative Verification

@@

  abstract interpretation--based invariant

@@

  analysis works precisely and efficiently for challenging loop and

@@

To the best of our knowledge, this article presents the first resource
analysis framework for C that makes it possible to combine non-trivial
automatically derived bounds with interactively derived bounds in a
proof system that produces verifiable certificates for the bounds.
Our approach complements existing work since it provides a semantic
foundation for the computation of bounds while still providing support
for automation that sometimes goes beyond the capabilities of existing
techniques.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we directly derive global bounds that are
functions of the input parameters of the program.  Another unique
feature of our system is, that it can handle resources like memory
that may become available during the execution.

Our starting point is a recent work~\cite{veristack14} in which we
have formally verified bounds on the stack usage of C programs.  A key
part of this stack-bound verification is based on a quantitative Hoare
logic for deriving abstract stack bounds for CompCert Clight programs
that depend on the depth of (recursive) function calls.  % For
% non-recursive programs it is possible to compute derivations for such
% stack bounds in the quantitative Hoare logic automatically.  
% By compiling these bounds together with the C program, we have derived
% verified stack bounds for x86 assembly code.
%
Our \emph{first contribution} is a generalization of the quantitative
Hoare logic for CompCert Clight that enables us to derive resource
bounds that are parametric in the resource of interest.  % The
% quantitative logic is implemented in Coq and proved sound with respect
% to a small-step operational cost semantics for Clight.
% that counts a
% constant resource consumption for each atomic execution step.  The
% resource consumption at each step is parametric in a user-definable cost
% metric and can be negative to express the release of resources (as
% needed, e.g., to reason about memory usage).  Furthermore, a specific
% \emph{tick} function can be inserted in the source code to further
% customize the cost model.
In the implementation of the logic, we use
a shallow embedding in Coq which makes the logic very flexible.  There
is practically no limitation on the format of the derived resource
bounds since they can be any function that is definable in Coq.  The
derived bounds can also be parametric in a set of cost metrics or
specifically apply to one fixed cost metric.

The precision and expressivity of the quantitative Hoare logic provide
an ample foundation for reasoning about resource consumption of Clight
programs.  % The logic is naturally compositional and the derived
% \emph{quantitative Hoare triples} can be used to specify the resource
% consumption of library functions or to describe resource contracts
% between different functions. 
However, reasoning about quantitative
properties can often be more tedious than reasoning about intentional
properties.  Consequently, automation is inevitable to derive resource
bounds for large code bases like CertiKOS.  Such an automation was
easily achievable in the case of constant stack bounds for a code base
that does not contain recursive functions~\cite{veristack14}.  In
general however, resource bounds have to be parametric in the
arguments of a function and depend on the number of loop iterations
and (recursive) function calls that are performed by the function.

For functional programs, there exist resource analysis systems---based
on type-based amortized resource analysis~\cite{Jost03,
  HoffmannAH12}---that can automatically derive complex polynomial
bounds with tight constant factors.  This type-based amortized
resource analysis has been an inspiration in the design of the
quantitative Hoare logic and is therefore a natural candidate to
automate the reasoning.  However, it is a well-known problem that
type-based amortized resource analysis works well for functional
programs with pattern matching but not for C-like programs with
control flow that depends integer arithmetic.  For example, it is a
long-time open problem how to use type-based amortized analysis to
derive bounds for functions whose resource consumption depends on
(possibly negative) integers and non-sequential control flow as
introduced by \code{break} and \code{return} statements.
Unfortunately, these are the functions with non-constant resource
consumption that are most common in CertiKOS and other system
software.

Our \emph{main contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive functions as
well as \code{break} and \code{return} statements.  Using
potential-based approach of amortized analysis allows us to
automatically handle programs that are need special treatment in other
techniques such as
\begin{itemize}
\item interaction between \emph{sequential loops or function calls} through
  size changes of variables,
\item \emph{nested loops} that influence each others number of iterations,
\item and \emph{tricky iteration patterns} as found for instance in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
%
The main innovation that makes the analysis practical for C is the use
of interval sizes in potential functions instead of the sizes of
variables.  This leads to precise bounds of programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract interpretation--based methods such as disjunctive invariant
generation~\cite{GulwaniMC09} and symbolic backward
execution~\cite{GulwaniZ10}.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

@@
